### 2020-09-04 submission.tex
Final touches

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended
% to minimize problems and delays during our production
% process. Please follow the template instructionssum 
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% Once your paper is accepted for publication,
% PLEASE REMOVE ALL TRACKED CHANGES in this file
% and leave only the final text of your manuscript.
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file.
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission.
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column.
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2".
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

@@

%Molecular clock models are used for dating evolutionary divergence times. 
Uncorrelated relaxed clock models enable estimation of molecular substitution rates across lineages and are widely used in phylogenetics for dating evolutionary divergence times.
In this article we delved into the internal complexities of the relaxed clock model in order to develop efficient MCMC operators for Bayesian phylogenetic inference.
We compared three substitution rate parameterisations,  introduced an adaptive operator which learns the weights of other operators during MCMC and, we explored how relaxed clock model estimation can benefit from two cutting-edge proposal kernels: the AVMVN and Bactrian kernels.
This work has produced an operator scheme that is up to 65 times more efficient at exploring continuous relaxed clock parameters compared with previous setups, depending on the dataset. %RB: 65 or 66 mentioned later?
Finally, we explored variants of the standard narrow exchange operator which are specifically designed for the relaxed clock model.
In the most extreme case, this new operator traversed tree space 40\% more efficiently than narrow exchange.
The methodologies introduced are adaptive and highly effective on short as well as long alignments.
The results are available via the open source optimised relaxed clock (ORC) package for BEAST 2 under the LGPL 2(?) licence (\url{https://github.com/jordandouglas/ORC}).

@@

Biological sequences, such as DNA, accumulate mutations over generations. 
By comparing such sequences in a phylogenetic framework, the evolutionary tree of lifeforms can be inferred.
With the overwhelming availability of biological sequence data, and the increasing affordability of collecting new data, the development of fast and efficient phylogenetic algorithms is more important than ever.
In this article we focus on the relaxed clock model, which is very popular in phylogenetics.
We explored how a range of optimisations can improve the statistical inference of the relaxed clock.
This work has produced a phylogenetic setup which can infer parameters related to the relaxed clock up to 65 times faster than previous setups, depending on the dataset. %RB: 65 or 66 mentioned later?
The methods introduced adapt to the dataset during computation and are highly efficient when processing long biological sequences.

@@

The simplest phylogenetic clock model -- the strict clock -- makes the mathematically convenient assumption that the evolutionary rate is constant across all lineages \cite{zuckerkandl1965evolutionary, kuhner1995estimating, larget1999markov}.  % Early examples? 
However, molecular substitution rates are known to vary over time, over population sizes, over evolutionary pressures, and over nucleic acid replicative machineries \cite{gillespie1994causes, woolfit2009effective, loh2010optimization}.
Moreover, any given dataset could be clock-like (where substitution rates have a small variance across lineages) or non clock-like (a large variance). 
In the latter case, a strict clock is probably not suitable. 
In the remainder of this paper %RB: "article" and "paper" are used a few times -- should we keep things consistent to either "paper" or "article"?
we will only consider uncorrelated relaxed clock models.

This led to the development of relaxed (uncorrelated) clock models, under which each branch in the phylogenetic tree has its own molecular substitution rate  \cite{drummond2006relaxed}.
Branch rates can be drawn from a range of probability distributions including log-normal, exponential, gamma, and inverse-gamma distributions \cite{drummond2006relaxed, lepage2007general, li2012model}.
This class of models is widely used, and has aided insight into many recent biological problems, including the 2016 Zika virus outbreak \cite{faria2017establishment} and the COVID-19 pandemic \cite{giovanetti2020first}.
% is compatible with increasingly prominent multispecies coalescent models \cite{ogilvie2017starBEAST 2},

@@

Let $\mathcal{T}$ be a binary rooted time tree with $N$ taxa, and %. Let $L$ be the number of sites within the multiple sequence alignment $D$. 
$D$ data associated with the tips, like a multiple sequence alignment with $L$ sites, morphological data or geographic locations. 
The posterior density of a phylogenetic model is described by

\begin{eqnarray}
\label{eq:bayesian}
p(\mathcal{T}, \vec{\mathcal{R}}^{\,}, \sigma, \mu_C, \theta|D) \propto  p(D|\mathcal{T}, r(\vec{\mathcal{R}}^{\,}), \mu_C, \theta) \; p(\mathcal{T}|\theta) \;  p(\vec{\mathcal{R}}^{\,} | \sigma) \; p(\sigma) \; p(\mu_C) \; p(\theta).
\end{eqnarray}

@@

\begin{eqnarray}
\label{eq:MCMC}
\alpha(x^\prime|x) =  \min\left( 1, \frac{p(x^\prime|D)}{p(x|D)} \; \frac{q(x|x^\prime)}{q(x^\prime|x)} \; |J| \right).
\end{eqnarray}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/rateparameterisation.pdf}
\caption{\textbf{Branch rate parameterisations.} Top left: the prior density of a branch rate $r$ under a Log-Normal($-0.5\sigma^2, \sigma$) distribution (with its mean fixed at 1). 
The function for transforming $\mathcal{R}$ into branch rates $r(\mathcal{R})$ is depicted for \textit{real} (top right), \textit{cat} (bottom left), and \textit{quant} (bottom right).
For visualisation purposes, there are only 10 bins/pieces displayed, however in practice we use $2N-2$ bins for \textit{cat} and 100 pieces for \textit{quant}.
The first and final \textit{quant} pieces are equal to the underlying function however the pieces in between use linear approximations of this function. }
\label{fig:rateparams}
\end{figure}

@@

The key advantage of the \textit{cat} parameterisation is the removal of a term from the posterior density (\textbf{Equation \ref{eq:bayesian}}), or more accurately the replacement of a non-trivial $p(\vec{\mathcal{R}}^{\,} | \sigma)$ term with that of a uniform prior. 
This may facilitate efficient exploration of the posterior distribution by MCMC.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/correlations.pdf}
\caption{\textbf{Traversing likelihood space.}
The z-axes above are the log-likelihoods of the genetic distance $r \times \tau$ between two simulated nucleic acid sequences of length $L$, under the Jukes-Cantor substitution model \cite{jukes1969evolution}. 
Two possible proposals from the current state (white circle) are depicted.
These proposals are generated by the \texttt{RandomWalk} (\texttt{RW}) and \texttt{ConstantDistance} (\texttt{CD}) operators.
In the low signal dataset ($L=0.1$kb), both operators can traverse the likelihood space effectively.
 However, the exact same proposal by \texttt{RandomWalk} incurs a much larger likelihood penalty in the $L=0.5$kb dataset by ``falling off the ridge'', in contrast to \texttt{ConstantDistance} which ``walks along the ridge''.
 This discrepancy is even stronger for larger datasets and thus necessitates the use of operators such as \texttt{ConstantDistance} which account for correlations between branch lengths and rates. }
\label{fig:landscape}
\end{figure}

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth]{Figures/clockSD.pdf}
\caption{\textbf{Clock standard deviation scale operators.}
The two operators above propose a clock standard deviation $\sigma \rightarrow \sigma^\prime$. 
Then, either the new quantiles are such that the rates remain constant (``New quantiles'', above) or the new rates are  such that the quantiles remain constant (``New rates''). 
In the \textit{real} parameterisation, these two operators are known as \texttt{Scale} and \texttt{CisScale}, respectively.
Whereas, in \textit{quant}, they are known as \texttt{CisScale} and \texttt{Scale}.    }
\label{fig:clockSDoperators}
\end{figure}

@@

It is not always clear which operators weighting scheme is best for a given dataset.
In this article we introduce \texttt{AdaptiveOperatorSampler} -- a meta-operator which learns the weights of other operators during MCMC and then samples these operators according to their learned weights.
This meta-operator undergoes three phases.
In the first phase (burn-in), \texttt{AdaptiveOperatorSampler} samples from its set of sub-operators uniformly at random.
In the second phase (learn-in), the meta-operator continues to sample operators uniformly at random however it begins learning several terms detailed below.
In its final phase, \texttt{AdaptiveOperatorSampler} samples operators (denoted by $\omega$) using the following distribution:

@@

\noindent
where  $\mathbb{T}(\omega)$ the cumulative computational time spent on each operator, $\mathbb{D}(.)$ a distance function and we use $\Omega = 0.01$ to allow any sub-operator to be sampled regardless of its performance.
The parameters of interest (POI) may be either a set of numerical parameters (such as branch rates or node heights), or it may be the tree itself, but it cannot be both in its current form.
The distance between state $x_p$ and its (accepted) proposal $x_p^\prime$ with respect to parameter $p$ is determined by

@@

Under \textbf{Equations \ref{eqn:adaptiveSampler} and \ref{eqn:distanceFunctions}}, operators which effect larger changes on the parameters of interest, in shorter runtime, are sampled with greater probabilities. 
Division of the squared distance by a parameters variance $\sigma^2_p$ enables comparison between different numerical parameters.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/bactrian.pdf}
\caption{\textbf{The Bactrian proposal kernel.} The step size made under a Bactrian proposal kernel is equal to $s \Sigma$ where $\Sigma$ is drawn from the above distribution and $s$ is tunable.   }
\label{fig:bactrian}
\end{figure}

@@

 & Operator(s) & Proposal & Parameter $x$   \\
   \hline
 1 & \texttt{RandomWalk}  & $x^\prime \leftarrow x + s\Sigma$ & $\vec{\mathcal{R}}^{\,}, \sigma$  \\
  \hline
 2 & \texttt{Scale} & $x^\prime \leftarrow x \times e^{s\Sigma}$ & $\vec{\mathcal{R}}^{\,}, \sigma$   \\
  \hline
 3 & \texttt{Interval} & $\begin{array} {rl} &y \leftarrow \frac{1 - x}{x} \times e^{s\Sigma} \\ &x^\prime \leftarrow \frac{y}{y + 1}  \end{array}$ & $\vec{\mathcal{R}}^{\,}$  \\
  \hline
 4 & \texttt{ConstantDistance} \texttt{SimpleDistance} & $x^\prime \leftarrow x + s\Sigma$ & $t$ \\
 \hline
 5 & \texttt{SmallPulley} & $x^\prime \leftarrow x + s\Sigma$ & $\vec{\mathcal{R}}^{\,}$  \\
 \hline
6 & \texttt{CisScale}  & $x^\prime \leftarrow x \times e^{s\Sigma}$ & $\sigma$   \\
\end{tabular}
\caption{Proposal kernels $q(x^\prime|x)$ of clock model operators.
 In each operator, $\Sigma$ is drawn from either a $\text{Bactrian}(m)$ or $\text{Uniform}$ distribution.  %RB: Uniform on which interval? (-1,1)?
 The scale size $s$ is tunable.
 \texttt{ConstantDistance} and \texttt{SimpleDistance} propose tree heights $t$.
  The \texttt{Interval} operator applies to rate quantiles and respects its domain i.e. $0 < x, x^\prime < 1$. }
\label{table:bactriankernels}
\end{table}

@@

% BRIEFLY describe three operators: random walk, scale, uniform + where the proposal kernal comes in + which parameters in the clock model it applies to

@@

The \texttt{NarrowExchange} operator \cite{drummond2002estimating}, used widely in BEAST \cite{drummond2012bayesian,suchard2018bayesian} and BEAST 2 \cite{bouckaert2019beast}, is similar to nearest-neighbour-interchange \cite{semple2003phylogenetics}, and works as follows (\textbf{Fig \ref{fig:narrowexchange}}):

@@

\textul{\textit{Step 2}}. Identify the child of $E$ with the greater height. Denote this child as $D$ and its sibling as $C$ (i.e. $t_D > t_C$).
 %RB: C must not be internal, right? If C is leaf, abort?

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/NarrowExchange.pdf}
\caption{\textbf{Depiction of \texttt{NarrowExchange} and \texttt{NarrowExchangeRate} operators.} Proposals are denoted by $\mathcal{T} \rightarrow \mathcal{T}^\prime$. The vertical axis corresponds to node height $t$. In the bottom figure, branch rates $r$ are indicated by line width and therefore genetic distances are equal to the width of each branch multiplied by its length. 
In this example, the $\mathcal{D}_{AE}$ and $\mathcal{D}_{CE}$ constraints are satisfied.
%RB: at this stage, D_AE and D_CE constraints have not been defined yet. Remove or reformulate or move to after definition?
}
\label{fig:narrowexchange}
\end{figure}

@@

Here, we present the \texttt{NarrowExchangeRate} (NER) operator. 
Let $r_A$, $r_B$, $r_C$, and $r_D$ be the substitution rates of nodes $A$, $B$, $C$, and $D$, respectively. 
In addition to the modest topological change applied by \texttt{NarrowExchange}, NER also proposes new branch rates ${r_A}^\prime$, ${r_B}^\prime$, ${r_C}^\prime$, and ${r_D}^\prime$. While NER does not alter $t_D$ (i.e. ${t_D}^\prime \leftarrow t_D$), we also consider NERw - a special case of the NER operator which embarks $t_D$ on a random walk:

@@

\noindent
for random walk step size $s\Sigma$ where $s$ is tunable and $\Sigma$ is drawn from a Uniform or Bactrian distribution. NER (and NERw) are compatible with both the \textit{real} and \textit{quant} parameterisations. 
Analogous to the \texttt{ConstantDistance} operator, the proposed rates ensure that the genetic distances between nodes $A$, $B$, $C$, and $E$ are constant. 
Let $\mathcal{D}_{ij}$ be the constraint defined by a constant genetic distance between nodes $i$ and $j$ before and after the proposal.
There are six pairwise distances between these four nodes and therefore there are six such constraints:

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/acceptanceRates.pdf}
\caption{\textbf{Screening of NER and NERw variants by acceptance rate.} Top left: comparison of NER variants with the null operator $\text{NER}\{\}$ = \texttt{NarrowExchange}. 
Each operator is represented by a single point, uniquely encoded by the point stylings. 
The number of times each operator is proposed and accepted is compared with that of $\text{NER}\{\}$, and one-sided z-tests are performed to assess the statistical significance between the two acceptance rates ($p = 0.001$).  This process is repeated across $300$ simulated datasets. The axes of each plot are the proportion of these $300$ simulations for which there is evidence that the operator is significantly better than $\text{NER}\{\}$ (x-axis) or worse than $\text{NER}\{\}$ (y-axis). Top right: comparison of NER and NERw acceptance rates. Each point is one NER/NERw variant from a single simulation. Bottom: relationship between the acceptance rates $\alpha$ of $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$ and $\text{NER}\{ \}$ with the clock model standard deviation $\sigma$ and the number of sites $L$.  Each point is a single simulation. }
\label{fig:acceptanceRateScreening}
\end{figure}

>>>>

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended
% to minimize problems and delays during our production
% process. Please follow the template instructions 
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% Once your paper is accepted for publication,
% PLEASE REMOVE ALL TRACKED CHANGES in this file
% and leave only the final text of your manuscript.
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file.
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission.
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column.
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2".
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

@@

%Molecular clock models are used for dating evolutionary divergence times. 
Uncorrelated relaxed clock models enable estimation of molecular substitution rates across lineages and are widely used in phylogenetics for dating evolutionary divergence times.
In this article we delved into the internal complexities of the relaxed clock model in order to develop efficient MCMC operators for Bayesian phylogenetic inference.
We compared three substitution rate parameterisations, introduced an adaptive operator which learns the weights of other operators during MCMC, and we explored how relaxed clock model estimation can benefit from two cutting-edge proposal kernels: the AVMVN and Bactrian kernels.
This work has produced an operator scheme that is up to 65 times more efficient at exploring continuous relaxed clock parameters compared with previous setups, depending on the dataset. 
Finally, we explored variants of the standard narrow exchange operator which are specifically designed for the relaxed clock model.
In the most extreme case, this new operator traversed tree space 40\% more efficiently than narrow exchange.
The methodologies introduced are adaptive and highly effective on short as well as long alignments.
The results are available via the open source optimised relaxed clock (ORC) package for BEAST 2 under a GNU licence (\url{https://github.com/jordandouglas/ORC}).

@@

Biological sequences, such as DNA, accumulate mutations over generations. 
By comparing such sequences in a phylogenetic framework, the evolutionary tree of lifeforms can be inferred.
With the overwhelming availability of biological sequence data, and the increasing affordability of collecting new data, the development of fast and efficient phylogenetic algorithms is more important than ever.
In this article we focus on the relaxed clock model, which is very popular in phylogenetics.
We explored how a range of optimisations can improve the statistical inference of the relaxed clock.
This work has produced a phylogenetic setup which can infer parameters related to the relaxed clock up to 65 times faster than previous setups, depending on the dataset. 
The methods introduced adapt to the dataset during computation and are highly efficient when processing long biological sequences.

@@

The simplest phylogenetic clock model -- the strict clock -- makes the mathematically convenient assumption that the evolutionary rate is constant across all lineages \cite{zuckerkandl1965evolutionary, kuhner1995estimating, larget1999markov}.  % Early examples? 
However, molecular substitution rates are known to vary over time, over population sizes, over evolutionary pressures, and over nucleic acid replicative machineries \cite{gillespie1994causes, woolfit2009effective, loh2010optimization}.
Moreover, any given dataset could be clock-like (where substitution rates have a small variance across lineages) or non clock-like (a large variance). 
In the latter case, a strict clock is probably not suitable.

This led to the development of relaxed (uncorrelated) clock models, under which each branch in the phylogenetic tree has its own molecular substitution rate  \cite{drummond2006relaxed}.
Branch rates can be drawn from a range of probability distributions including log-normal, exponential, gamma, and inverse-gamma distributions \cite{drummond2006relaxed, lepage2007general, li2012model}.
This class of models is widely used, and has aided insight into many recent biological problems, including the 2016 Zika virus outbreak \cite{faria2017establishment} and the COVID-19 pandemic \cite{giovanetti2020first}.
In the remainder of this paper we will only consider uncorrelated relaxed clock models.
% is compatible with increasingly prominent multispecies coalescent models \cite{ogilvie2017starBEAST 2},

@@

Let $\mathcal{T}$ be a binary rooted time tree with $N$ taxa, and %. Let $L$ be the number of sites within the multiple sequence alignment $D$. 
data $D$ associated with the tips, such as a multiple sequence alignment with $L$ sites, morphological data or geographic locations. 
The posterior density of a phylogenetic model is described by

\begin{eqnarray}
\label{eq:bayesian}
p(\mathcal{T}, \vec{\mathcal{R}}^{\,}, \sigma, \mu_C, \theta|D) \propto  p(D|\mathcal{T}, r(\vec{\mathcal{R}}^{\,}), \mu_C, \theta) \; p(\mathcal{T}|\theta) \;  p(\vec{\mathcal{R}}^{\,} | \sigma) \; p(\sigma) \; p(\mu_C) \; p(\theta)
\end{eqnarray}

@@

\begin{eqnarray}
\label{eq:MCMC}
\alpha(x^\prime|x) =  \min\left( 1, \frac{p(x^\prime|D)}{p(x|D)} \; \frac{q(x|x^\prime)}{q(x^\prime|x)} \; |J| \right)
\end{eqnarray}

@@

\begin{figure}[!h]
%%%%% \includegraphics[width=\textwidth]{Figures/rateparameterisation.pdf}
\caption{\textbf{Branch rate parameterisations.} Top left: the prior density of a branch rate $r$ under a Log-Normal($-0.5\sigma^2, \sigma$) distribution (with its mean fixed at 1). 
The function for transforming $\mathcal{R}$ into branch rates $r(\mathcal{R})$ is depicted for \textit{real} (top right), \textit{cat} (bottom left), and \textit{quant} (bottom right).
For visualisation purposes, there are only 10 bins/pieces displayed, however in practice we use $2N-2$ bins for \textit{cat} and 100 pieces for \textit{quant}.
The first and final \textit{quant} pieces are equal to the underlying function however the pieces in between use linear approximations of this function. }
\label{fig:rateparams}
\end{figure}

@@

The key advantage of the \textit{cat} parameterisation is the removal of a term from the posterior density (\textbf{Eq \ref{eq:bayesian}}), or more accurately the replacement of a non-trivial $p(\vec{\mathcal{R}}^{\,} | \sigma)$ term with that of a uniform prior. 
This may facilitate efficient exploration of the posterior distribution by MCMC.

@@

\begin{figure}[!h]
%%%%% \includegraphics[width=\textwidth]{Figures/correlations.pdf}
\caption{\textbf{Traversing likelihood space.}
The z-axes above are the log-likelihoods of the genetic distance $r \times \tau$ between two simulated nucleic acid sequences of length $L$, under the Jukes-Cantor substitution model \cite{jukes1969evolution}. 
Two possible proposals from the current state (white circle) are depicted.
These proposals are generated by the \texttt{RandomWalk} (\texttt{RW}) and \texttt{ConstantDistance} (\texttt{CD}) operators.
In the low signal dataset ($L=0.1$kb), both operators can traverse the likelihood space effectively.
 However, the exact same proposal by \texttt{RandomWalk} incurs a much larger likelihood penalty in the $L=0.5$kb dataset by ``falling off the ridge'', in contrast to \texttt{ConstantDistance} which ``walks along the ridge''.
 This discrepancy is even stronger for larger datasets and thus necessitates the use of operators such as \texttt{ConstantDistance} which account for correlations between branch lengths and rates. }
\label{fig:landscape}
\end{figure}

@@

\begin{figure}[!h]
\centering
%%%%% \includegraphics[width=\textwidth]{Figures/clockSD.pdf}
\caption{\textbf{Clock standard deviation scale operators.}
The two operators above propose a clock standard deviation $\sigma \rightarrow \sigma^\prime$. 
Then, either the new quantiles are such that the rates remain constant (``New quantiles'', above) or the new rates are  such that the quantiles remain constant (``New rates''). 
In the \textit{real} parameterisation, these two operators are known as \texttt{Scale} and \texttt{CisScale}, respectively.
Whereas, in \textit{quant}, they are known as \texttt{CisScale} and \texttt{Scale}.    }
\label{fig:clockSDoperators}
\end{figure}

@@

It is not always clear which operator weighting scheme is best for a given dataset.
In this article we introduce \texttt{AdaptiveOperatorSampler} -- a meta-operator which learns the weights of other operators during MCMC and then samples these operators according to their learned weights.
This meta-operator undergoes three phases.
In the first phase (burn-in), \texttt{AdaptiveOperatorSampler} samples from its set of sub-operators uniformly at random.
In the second phase (learn-in), the meta-operator starts learning several terms detailed below whilst continuing to sample operators uniformly at random.
In its final phase, \texttt{AdaptiveOperatorSampler} samples operators (denoted by $\omega$) using the following distribution:

@@

\noindent
where $\mathbb{T}(\omega)$ is the cumulative computational time spent on each operator, $\mathbb{D}$ is a distance function, and we use $\Omega = 0.01$ to allow any sub-operator to be sampled regardless of its performance.
The parameters of interest (POI) may be either a set of numerical parameters (such as branch rates or node heights), or it may be the tree itself, but it cannot be both in its current form.
The distance between state $x_p$ and its (accepted) proposal $x_p^\prime$ with respect to parameter $p$ is determined by

@@

Under \textbf{Equations \ref{eqn:adaptiveSampler} and \ref{eqn:distanceFunctions}}, operators which effect larger changes on the parameters of interest, in shorter runtime, are sampled with greater probabilities. 
Division of the squared distance by a parameter's sample variance $\sigma^2_p$ enables comparison between different numerical parameters.

@@

\begin{figure}[!h]
%%%%% \includegraphics[width=\textwidth]{Figures/bactrian.pdf}
\caption{\textbf{The Bactrian proposal kernel.} The step size made under a Bactrian proposal kernel is equal to $s \Sigma$ where $\Sigma$ is drawn from the above distribution and $s$ is tunable.   }
\label{fig:bactrian}
\end{figure}

@@

 & Operator(s) & Proposal & Parameter $x$   \\
   \hline
 1 & \texttt{RandomWalk}  & $x^\prime \leftarrow x + s\Sigma$ & $\vec{\mathcal{R}}^{\,}, \sigma$  \\
  \hline
 2 & \texttt{Scale} & $x^\prime \leftarrow x \times e^{s\Sigma}$ & $\vec{\mathcal{R}}^{\,}, \sigma$   \\
  \hline
 3 & \texttt{Interval} & $\begin{array} {rl} &y \leftarrow \frac{1 - x}{x} \times e^{s\Sigma} \\ &x^\prime \leftarrow \frac{y}{y + 1}  \end{array}$ & $\vec{\mathcal{R}}^{\,}$  \\
  \hline
 4 & \texttt{ConstantDistance} \texttt{SimpleDistance} & $x^\prime \leftarrow x + s\Sigma$ & $t$ \\
 \hline
 5 & \texttt{SmallPulley} & $x^\prime \leftarrow x + s\Sigma$ & $\vec{\mathcal{R}}^{\,}$  \\
 \hline
6 & \texttt{CisScale}  & $x^\prime \leftarrow x \times e^{s\Sigma}$ & $\sigma$   \\
\end{tabular}
\caption{Proposal kernels $q(x^\prime|x)$ of clock model operators.
 In each operator, $\Sigma$ is drawn from either a $\text{Bactrian}(m)$ or $\text{uniform}$ distribution.  %RB: Uniform on which interval? (-1,1)?
 The scale size $s$ is tunable.
 \texttt{ConstantDistance} and \texttt{SimpleDistance} propose tree heights $t$.
  The \texttt{Interval} operator applies to rate quantiles and respects its domain i.e. $0 < x, x^\prime < 1$. }
\label{table:bactriankernels}
\end{table}

@@

% BRIEFLY describe three operators: random walk, scale, uniform + where the proposal kernel comes in + which parameters in the clock model it applies to

@@

The \texttt{NarrowExchange} operator \cite{drummond2002estimating}, used widely in BEAST \cite{drummond2012bayesian,suchard2018bayesian} and BEAST 2 \cite{bouckaert2019beast}, is similar to nearest neighbour interchange \cite{semple2003phylogenetics}, and works as follows (\textbf{Fig \ref{fig:narrowexchange}}):

@@

\textul{\textit{Step 2}}. Identify the child of $E$ with the greater height. Denote this child as $D$ and its sibling as $C$ (i.e. $t_D > t_C$). If $D$ is a leaf node, then reject the proposal.

@@

\begin{figure}[!h]
%%%%% \includegraphics[width=\textwidth]{Figures/NarrowExchange.pdf}
\caption{\textbf{Depiction of \texttt{NarrowExchange} and \texttt{NarrowExchangeRate} operators.} Proposals are denoted by $\mathcal{T} \rightarrow \mathcal{T}^\prime$. The vertical axes correspond to node heights $t$. In the bottom figure, branch rates $r$ are indicated by line width and therefore genetic distances are equal to the width of each branch multiplied by its length. 
In this example, the $\mathcal{D}_{AE}$ and $\mathcal{D}_{CE}$ constraints are satisfied.
%RB: at this stage, D_AE and D_CE constraints have not been defined yet. Remove or reformulate or move to after definition?
}
\label{fig:narrowexchange}
\end{figure}

@@

Here, we present the \texttt{NarrowExchangeRate} (NER) operator. 
Let $r_A$, $r_B$, $r_C$, and $r_D$ be the substitution rates of nodes $A$, $B$, $C$, and $D$, respectively. 
In addition to the modest topological change applied by \texttt{NarrowExchange}, NER also proposes new branch rates ${r_A}^\prime$, ${r_B}^\prime$, ${r_C}^\prime$, and ${r_D}^\prime$. While NER does not alter $t_D$ (i.e. ${t_D}^\prime \leftarrow t_D$), we also consider NERw -- a special case of the NER operator which embarks $t_D$ on a random walk:

@@

\noindent
for random walk step size $s\Sigma$ where $s$ is tunable and $\Sigma$ is drawn from a uniform or Bactrian distribution. NER (and NERw) are compatible with both the \textit{real} and \textit{quant} parameterisations. 
Analogous to the \texttt{ConstantDistance} operator, the proposed rates ensure that the genetic distances between nodes $A$, $B$, $C$, and $E$ are constant. 
Let $\mathcal{D}_{ij}$ be the constraint defined by a constant genetic distance between nodes $i$ and $j$ before and after the proposal.
There are six pairwise distances between these four nodes and therefore there are six such constraints:

@@

\begin{figure}[!h]
%%%%% \includegraphics[width=\textwidth]{Figures/acceptanceRates.pdf}
\caption{\textbf{Screening of NER and NERw variants by acceptance rate.} Top left: comparison of NER variants with the null operator $\text{NER}\{\}$ = \texttt{NarrowExchange}. 
Each operator is represented by a single point, uniquely encoded by the point stylings. 
The number of times each operator is proposed and accepted is compared with that of $\text{NER}\{\}$, and one-sided z-tests are performed to assess the statistical significance between the two acceptance rates ($p = 0.001$).  This process is repeated across $300$ simulated datasets. The axes of each plot are the proportion of these $300$ simulations for which there is evidence that the operator is significantly better than $\text{NER}\{\}$ (x-axis) or worse than $\text{NER}\{\}$ (y-axis). Top right: comparison of NER and NERw acceptance rates. Each point is one NER/NERw variant from a single simulation. Bottom: relationship between the acceptance rates $\alpha$ of $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$ and $\text{NER}\{ \}$ with the clock model standard deviation $\sigma$ and the number of sites $L$.  Each point is a single simulation. }
\label{fig:acceptanceRateScreening}
\end{figure}
